A Container-Based Cloud-Native Architecture for the Reproducible Execution of Multi-population Optimization Algorithms

### 2020-10-16 draft.tex
Grammar correction according to GDrive write refs #27

The trend concerning the parallel execution multi-population algorithms goes from
earlier hardware-based implementations using transputers
\cite{gorges1990explicit} through 
multi-core systems \cite{Serrano2008,lai2019adaptive} to multi-threaded \cite{merelo2019scaling} ones.
Moreover, recently, the focus has been on exploiting a higher number of processing
units by using GPUs \cite{tan2015survey,li2007efficient}, or distributed
systems, including web-based \cite{JSON} systems,
map-reduce \cite{fazenda2012} implementations,  grids \cite{munawar2010design,Gonzalez09},
voluntary computing systems \cite{MilkyWay,merelo2016nodio}, 
and, more importantly, cloud computing
\cite{GValdez2015,salza2019speed,valenzuela2015implementing,FlexGP}.

%% JJ Can you improve the redability of the paragraph below? 
%% Grammarly suggests that is hard to read, verly long sentences.
%% Edited a little - Mario
Cloud computing has become the standard way of running
enterprise applications.  Not only because of the convenience of the
pay-as-you-go model or the non-existent sunk cost of physical facilities, but also because
it offers a way of describing the infrastructure as part of the code, so that it
is much easier to reproduce results and this has been a boon for scientific
computing.  
However,  cloud computing has also been evolving, going from simply
putting old-style monolithic applications on virtual machines executing on an external 
data center to microservices \cite{microservices} and serverless architectures \cite{varghese2018next,Varghese2018849} 
that favor the parallel and asynchronous communication of
heterogeneous resources. In the process, a new methodology for designing 
{\em cloud-native applications} has been created, with brand new methods and
techniques \cite{Baldini2016287}. 
In these architectures, services or even functions are seen as independent 
processing nodes, departing from the monolithic
and even distributed paradigms, and becoming a loosely coupled collection of {\em
stateless functions} \cite{malawski2017serverless}, that react to events. 
%and only come into existence while they are doing some processing.
Systems developed with the patterns outlined above have the
properties of a reactive system \cite{boner2014reactive}, and they are generally 
more flexible, fault-tolerant, and scalable. % Some properties and reference - Mario  
% Again, you have to justify the need for a reactive architecture and define it before this paragraph - JJ

@@

\subsection{Event Driven Model} 
\label{edm}
Based on a reactive architecture we proposed in a previous work
\cite{guervos2018introducing}, we now describe the general architecture shown in
Figure \ref{fig:kafkEO}. Again, we can explain the model using the analogy
of producers and consumers of messages. First, we can see two queues, 
one labeled \texttt{Input} and the other one \texttt{Output}. In the diagram, 
push operations on a queue are represented by solid arrows connecting to the left side
of the queue box, and pull or pop operations as solid arrows leaving from the right side.
The architecture has at least four processes indicated in the diagram as
swimlanes: First, the \texttt{Setup} process, responsible for the reading a configuration
file and creating the initial populations. Second, the \texttt{Controller} process,
responsible for the migration between populations and keeping track of the algorithm's
iterations. The \texttt{Message Queue} process runs the \texttt{Input} and \texttt{Output}
queues. Finally, there is at least one \texttt{Stateless Function} process responsible
for running the isolated algorithms. In the example, two processes \texttt{PSO} and \texttt{GA},
are shown.

@@

Worker containers include a Python script called \texttt{main.py} running an
infinite loop, continually trying to pull a new population to work on it. Once
\texttt{main.py} receives a message containing configuration data, it creates a
worker object responsible for initializing and running the specified stateless
function (i.e., GA or PSO) with the population and configuration inside the message.
Once the function returns the evolved population, the
the main script pushes the results to the \texttt{Output} queue, and continues
to execute the infinite loop. This process is shown in Figure \ref{fig:worker}.

@@

In the first level of integration, developers
can implement a new stateless Docker container following
the description in the previous section (\ref{workers}). The worker
container executes a daemon that continuously pulls messages from the Redis host's input queue
specified as an environment variable {\texttt REDIS\_HOST}. If the new
population-based algorithm is written in Python, developers can base the new
implementation on the provided worker container image, adding a new algorithm
function and worker and modifying the \texttt{main.py} file. 
The worker must include a constructor with the configuration dictionary 
(see Listing \ref{code:config}) as the only parameter and call a the new function 
with the required parameters, unpacking them from the configuration dictionary if necessary.
All the code of the EvoSwarm Application is published with an open-source MIT license.

If developers use an existing library for population-based algorithms, they need
to change the standard model of execution, which creates a random
population because, in this case, the algorithm needs to start with the
population provided as a parameter. Depending on the evolutionary (or other) algorithm library used,
it may be required to extract the population from the
JSON message and replace the population object created by the default library implementation. For
example, the DEAP library we used for the GA algorithm creates the type of the
population at runtime (i.e., bit, float), so we have to replace the population
once DEAP created it. In the case of EvoloPy, the PSO method directly uses a
list of floats for storing the swarm, so we just added a parameter to receive
the population as a list of floats. Both libraries store information about each
iteration, so we just read it and put it in the format required.
Again, the population's state must be encoded as a JSON message to be
returned as a result, with additional data as
depicted in Listing \ref{code:result}. The output message contains worker container's id 
together with records giving information about each iteration
of the algorithm, including the best solution found. The controller uses 
this information to track the algorithm's execution and keep a 
log in the Redis container for posterior analyses.

@@

  \item[A container orchestration service] these services are compatible with {\tt docker-compose}
    files for configuration and provide transparent provisioning. Examples of these
    services are Amazon Elastic Container Service,  Red Hat Openshift
    (which is also open-source), and Marantis Docker Enterprise using Docker Swarm.
    Salza et al. \cite{salza2019speed} tested a variant of this option by using a CoreOs cluster.

@@

Because all these instances run in a private subnet, we also
run a Bastion instance, that is, a specially protected network node, in a public subnet, from where we send the
commands to execute the experiments
and later collected the results. We connected to this instance through a {\texttt ssh} session from 
our personal workstation. This is a {\texttt R5AD Large} instance, selected with 75GiB of storage, needed for
keeping the logs, and more memory (16GiB) for processing the logs in memory. The details these EC2 instances 
and their cost at the time of writing are shown in Table \ref{tab:ec2}.

@@

The complete configuration files are available at public GitHub repository 
\footnote{\url{https://github.com/mariosky/EvoSwarm/tree/master/aws-EC2}}, and
they are used on the next section to carry an extensive evaluation of the framework.
Again, each configuration describes the deployment of containers in a consistent
and portable way, an it was used in each phase of the life cycle, from
development through testing to production independently on the kind of
infrastructure.

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{eval_ratio}
  \caption{Boxplot of the number of evaluations per second, for 30 instances of function $f_4$,
          with different number of workers and number of populations. The $y$ axis is in log scale.}
  \label{fig:spworker:evalspersecond}
\end{figure}

@@

Table~\ref{tab:speedup-table} shows the speed-up obtained against the baseline of
one worker. We achieved the maximum speed-up when using four workers or more with
twenty populations. On the other hand, the maximum speed-up when we consider the
cost it was with eight workers. The best speed-ups for each number of workers are
shown in bold in the table.

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{evals_per_instance}
  \caption{Boxplot of the number of function evaluations required to reach the target, for 30 instances of function $f_4$, with a different number of workers and populations sizes}
  \label{fig:evals_instance}
\end{figure}

@@

A multi-population based algorithm can decrease the execution time
thanks to running the algorithm on several populations at the same
time. But, if the populations are heterogeneous 
it might enhance evolutionary search and avoid unneeded
evaluations to a greater extent than homogeneous systems systems do; heterogeneous settings, if
done right, increase the diversity of the whole population\cite{araujo2008multikulti}.

@@

In order to run this experiment, we just needed to edit a configuration file and
start it up. The plots presented in the results are generated by the BBOB
framework from the JSON log file for each run. There was no additional setup
needed, proving how this framework is able to not only offer an reproducible way
to run experiments, but also how easily it interfaces with other common tools in
the EC ecosystem.

@@

In this paper we have presented the EvoSwarm framework for 
the reproducible execution of experiments with an event-based, cloud-native architecture 
suitable for hosting multi-population, multi-algorithm optimization
methods. 
The framework is based in industry-standard development
tools such as Docker and Docker Compose, making it easy for scientists and practitioners to develop and deploy
% The last version still uses Docker Compose? - JJ
% Well it uses docker-compose files, and these files are  
% converted to tasks and services by the ECS-CLI. -Mario
% It still can be used to run the application locally - Mario 
multi-algorithm distributed experiments. The model implemented by the
framework is based on the asynchronous
exchange of messages between stateless functions that react to a continuous
stream of data; populations are the messages that flow in this stream. We show the differences of the message-driven architecture we presented against
four other proposals for multi-population algorithms found in the literature, and highlight the 
advantages for conducting asynchronous multi-population algorithms.

@@

This work also opens new possibilities. From the design point of view, it would
be interesting to use automatic scaling, instead of setting the number of
workers by hand. This would need a certain amount of configuration, as we would
be using container orchestration with auto-scaling capabilities such as
Kubernetes or Docker Swarm. The current orchestration (Amazon ECS) also has
auto-scaling capabilities, but additional configuration is needed; also, we need
to establish the auto-scaling of both containers and Amazons EC2 instances.
Instead of setting a fixed number of workers, only the maximum amount would need
to be established, which could be done directly or by setting an evaluation
budget.

>>>>

The trend concerning parallel execution multi-population algorithms goes from
earlier hardware-based implementations using transputers
\cite{gorges1990explicit} through 
multi-core systems \cite{Serrano2008,lai2019adaptive} to multi-threaded \cite{merelo2019scaling} ones.
Moreover, recently, the focus has been on exploiting a higher number of processing
units by using GPUs \cite{tan2015survey,li2007efficient}, or distributed
systems, including web-based \cite{JSON} systems,
map-reduce \cite{fazenda2012} implementations,  grids \cite{munawar2010design,Gonzalez09},
voluntary computing systems \cite{MilkyWay,merelo2016nodio}, 
and, more importantly, cloud computing
\cite{GValdez2015,salza2019speed,valenzuela2015implementing,FlexGP}.

%% JJ Can you improve the redability of the paragraph below? 
%% Grammarly suggests that is hard to read, verly long sentences.
%% Edited a little - Mario
Cloud computing has become the standard way of running
enterprise applications.  Not only because of the convenience of the
pay-as-you-go model or the non-existent sunk cost of physical facilities, but also because
it offers a way of describing the infrastructure as part of the code, so that it
is much easier to reproduce results and this has been a boon for scientific
computing.  
However,  cloud computing has also been evolving, going from simply
putting old-style monolithic applications on virtual machines executing on an external 
data center to microservices \cite{microservices} and serverless architectures \cite{varghese2018next,Varghese2018849} 
that favor the parallel and asynchronous communication of
heterogeneous resources. In the process, a new methodology for designing 
{\em cloud-native applications} has been created, with brand new methods and
techniques \cite{Baldini2016287}. 
In these architectures, services or even functions are seen as independent 
processing nodes, departing from the monolithic
and even distributed paradigms, and becoming a loosely coupled collection of {\em
stateless functions}\cite{malawski2017serverless}, that react to events. 
%and only come into existence while they are doing some processing.
Systems developed with the patterns outlined above have the
properties of a reactive system \cite{boner2014reactive}, and they are generally 
more flexible, fault-tolerant, and scalable. % Some properties and reference - Mario  
% Again, you have to justify the need for a reactive architecture and define it before this paragraph - JJ

@@

\subsection{Event Driven Model} 
\label{edm}
Based on a reactive architecture we proposed in a previous work
\cite{guervos2018introducing}, we now describe the general architecture shown in
Figure \ref{fig:kafkEO}. Again, we can explain the model using the analogy
of producers and consumers of messages. First, we can see two queues, 
one labeled \texttt{Input} and the other one \texttt{Output}. In the diagram, 
push operations on a queue are represented by solid arrows connecting to the left side
of the queue box, and pull or pop operations as solid arrows leaving from the right side.
The architecture has at least four processes indicated in the diagram as
swimlanes: First, the \texttt{Setup} process, responsible for reading a configuration
file and creating the initial populations. Second, the \texttt{Controller} process,
responsible for the migration between populations and keeping track of the algorithm's
iterations. The \texttt{Message Queue} process runs the \texttt{Input} and \texttt{Output}
queues. Finally, there is at least one \texttt{Stateless Function} process responsible
for running the isolated algorithms. In the example, two processes \texttt{PSO} and \texttt{GA},
are shown.

@@

Worker containers include a Python script called \texttt{main.py} running an
infinite loop, continually trying to pull a new population to work on it. Once
\texttt{main.py} receives a message containing configuration data, it creates a
worker object responsible for initializing and running the specified stateless
function (i.e., GA or PSO) with the population and configuration inside the message.
Once the function returns the evolved population,
the main script pushes the results to the \texttt{Output} queue, and continues
to execute the infinite loop. This process is shown in Figure \ref{fig:worker}.

@@

In the first level of integration, developers
can implement a new stateless Docker container following
the description in the previous section (\ref{workers}). The worker
container executes a daemon that continuously pulls messages from the Redis host's input queue
specified as an environment variable {\texttt REDIS\_HOST}. If the new
population-based algorithm is written in Python, developers can base the new
implementation on the provided worker container image, adding a new algorithm
function and worker and modifying the \texttt{main.py} file. 
The worker must include a constructor with the configuration dictionary 
(see Listing \ref{code:config}) as the only parameter and call the new function 
with the required parameters, unpacking them from the configuration dictionary if necessary.
All the code of the EvoSwarm Application is published with an open-source MIT license.

If developers use an existing library for population-based algorithms, they need
to change the standard model of execution, which creates a random
population because, in this case, the algorithm needs to start with the
population provided as a parameter. Depending on the evolutionary (or other) algorithm library used,
it may be required to extract the population from the
JSON message and replace the population object created by the default library implementation. For
example, the DEAP library we used for the GA algorithm creates the type of the
population at runtime (i.e., bit, float), so we have to replace the population
once DEAP creates it. In the case of EvoloPy, the PSO method directly uses a
list of floats for storing the swarm, so we just added a parameter to receive
the population as a list of floats. Both libraries store information about each
iteration, so we just read it and put it in the format required.
Again, the population's state must be encoded as a JSON message to be
returned as a result, with additional data as
depicted in Listing \ref{code:result}. The output message contains the
worker container's id 
together with records giving information about each iteration
of the algorithm, including the best solution found. The controller uses 
this information to track the algorithm's execution and keep a 
log in the Redis container for posterior analyses.

@@

  \item[A container orchestration service]: These services are compatible with {\tt docker-compose}
    files for configuration and provide transparent provisioning. Examples of these
    services are Amazon Elastic Container Service,  Red Hat Openshift
    (which is also open-source), and Marantis Docker Enterprise using Docker Swarm.
    Salza et al. \cite{salza2019speed} tested a variant of this option by using a CoreOs cluster.

@@

Because all these instances run in a private subnet, we also run a
Bastion instance, that is, a specially protected network node, in a
public subnet, from where we send the commands to execute the
experiments and later collect the results. We connected to this
instance through a {\texttt ssh} session from our personal
workstation. This is a {\texttt R5AD Large} instance, selected with
75GiB of storage, needed for keeping the logs, and more memory (16GiB)
for processing the logs in memory. The details these EC2 instances and
their cost at the time of writing are shown in Table \ref{tab:ec2}.

@@

The complete configuration files are available at public GitHub repository 
\footnote{\url{https://github.com/mariosky/EvoSwarm/tree/master/aws-EC2}}, and
they are used on the next section to carry an extensive evaluation of the framework.
Again, each configuration describes the deployment of containers in a consistent
and portable way, and it was used in each phase of the life cycle, from
development through testing to production independently on the kind of
infrastructure.

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{eval_ratio}
  \caption{Boxplot of the number of evaluations per second, for 30 instances of function $f_4$,
          with different number of workers and  populations. The $y$ axis is in log scale.}
  \label{fig:spworker:evalspersecond}
\end{figure}

@@

Table~\ref{tab:speedup-table} shows the speed-up obtained against the baseline of
one worker. We achieved the maximum speed-up when using four workers or more with
twenty populations. On the other hand, the maximum speed-up when we consider the
cost was with eight workers. The best speed-ups for each number of workers are
shown in bold in the table.

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{evals_per_instance}
  \caption{Boxplot of the number of function evaluations required to reach the target, for 30 instances of function $f_4$, with a different number of workers and population sizes}
  \label{fig:evals_instance}
\end{figure}

@@

A multi-population based algorithm can decrease the execution time
thanks to running the algorithm on several populations at the same
time. But, if the populations are heterogeneous 
it might enhance evolutionary search and avoid unneeded
evaluations to a greater extent than homogeneous systems do; heterogeneous settings, if
done right, increase the diversity of the whole population\cite{araujo2008multikulti}.

@@

In order to run this experiment, we just needed to edit a configuration file and
start it up. The plots presented in the results are generated by the BBOB
framework from the JSON log file for each run. There was no additional setup
needed, proving how this framework is able to not only offer a reproducible way
to run experiments, but also how easily it interfaces with other common tools in
the EC ecosystem.

@@

In this paper we have presented the EvoSwarm framework for the
reproducible execution of experiments with an event-based,
cloud-native architecture suitable for hosting multi-population,
multi-algorithm optimization methods.  The framework is based on
industry-standard development tools such as Docker and Docker Compose,
making it easy for scientists and practitioners to develop and deploy
multi-algorithm distributed experiments. The model implemented by the
framework is based on the asynchronous exchange of messages between
stateless functions that react to a continuous stream of data;
populations are the messages that flow in this stream. We show the
differences of the message-driven architecture we presented against
four other proposals for multi-population algorithms found in the
literature, and highlight the advantages for conducting asynchronous
multi-population algorithms.

@@

This work also opens new possibilities. From the design point of view, it would
be interesting to use automatic scaling, instead of setting the number of
workers by hand. This would need a certain amount of configuration, as we would
be using container orchestration with auto-scaling capabilities such as
Kubernetes or Docker Swarm. The current orchestration (Amazon ECS) also has
auto-scaling capabilities, but additional configuration is needed; also, we need
to establish the auto-scaling of both containers and Amazon's EC2 instances.
Instead of setting a fixed number of workers, only the maximum amount would need
to be established, which could be done directly or by setting an evaluation
budget.
